Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-16 manuscript/manuscript.tex
Finish Results.

\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations, however, do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.

@@

In contrast, the beta coalescent generates $\hilopmi$ that is consistently greater than the constant-$N$ Kingman.
As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model then in models of population growth than generate similar distortions in the SFS.
[IDEA: A plot of $\hilopmi$ at a fixed r vs Tajima's D for all the models together. Could make several plots, varying $r$ and $i_c$. This might be more convincing than the current figure.]
This is true across recombination rates and high/low cutoff minor allele counts.
These results demonstrate that $\hilopmi$ is capable of discriminating between coalescent models even when there is recombination between sites.

@@

Our results above show that $\fpmi$ and its coarse-grained analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data using any relevant method.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

@@

We can get an intuitive understanding for the result above by considering a sample four chromosomes.
In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
Genealogies with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
Conditional on observing a doubleton at a site, it is thus more likely that the genealogy has topology (B) and so the expected number of singletons at sites with the same genealogy is lower than average.
In terms of the 2-SFS, we have $\E{\eta_{12}} < \E{\eta_{1}} \E{\eta_{2}}$.

@@

As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
In the latter model, we vary both the fold-change in $N$ and the time of the change.
Like Tajima's D, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
Moreover, the two different models of growth collapse onto one another.
This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.

>>>>

\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

In contrast, the beta coalescent generates $\hilopmi$ that is consistently greater than the constant-$N$ Kingman.
As with non-recombining sites, the $\hilopmi$ is also greater in the beta coalescent model then in models of population growth than generate similar distortions in the SFS.
This is true across recombination rates and high/low cutoff minor allele counts.
These results demonstrate that $\hilopmi$ is capable of discriminating between coalescent models even when there is recombination between sites.

\Fig{fig:hilo_vs_tajD} shows the relative invariance of $\hilopmi$ to a time-varying coalescent rate.
While population growth in a Kingman coalescent model can strongly distort the SFS, as measured by Tajima's D, it has very little effect on the $\hilopmi$ for sites $r \E{T_2} \sim 1$ apart.
This holds for exponential as well as piecewise constant growth.
In contrast, the beta coalescent induces large positive $\hilopmi$ for the same values of Tajima's D.
This result is robust to the choice of cutoff minor allele count for binning.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/hilopmi_vs_tajimasD.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus normalized Tajima's D for simulations with recombination. All values are for $r \E{T_2} \sim 1$ and $n=100$. Orange points show Kingman coalescent simulations with population growth. Blue points show beta coalescent simulations. [Q: Should this be the bottom row of the previous figure?] \label{fig:hilo_vs_tajD}}
\end{figure}

@@

Our results above show that $\fpmi$ and its coarse-grained analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

@@

We can get an intuitive understanding for these results by considering a sample four chromosomes.
In the Kingman coalescent, there are only two possible tree topologies (\Fig{fig:trees}).
Furthermore, the total branch length is independent of the topology \cite{Wakeley2009}.
As a result, there is a trade-off between the length of branches leading to singleton/tripleton mutations on one hand, and the branch length leading to doubletons on the other.
Genealogies with topology (A) will have more opportunities for the former and loci with topology (B) will have more opportunities for the latter.
Conditional on observing a doubleton at a site, it is thus more likely that the genealogy has topology (B) and so the expected number of singletons at sites with the same genealogy is lower than average.
In terms of the 2-SFS, we have $\E{\eta_{12}} < \E{\eta_{1}} \E{\eta_{2}}$.

@@

As argued above, $\E{\eta_{12}}$ is also distorted by changes in the coalescent time distribution induced by population growth.
\Fig{fig:sdpmi_vs_sdratio} demonstrates that $\fpmi$ accounts for this fact by normalizing by the SFS.
\Fig{fig:sdpmi_vs_sdratio} plots $\fpmi_{12}$ against the ratio of singletons to doubletons $\eta_1 / \eta_2$ for the beta coalescent and two models of population growth: exponential growth and a piecewise-constant model with two epochs.
In the latter model, we vary both the fold-change in $N$ and the time of the change.
Like Tajima's D, the singleton/doubleton ratio measures distortion the SFS relative to the constant-$N$ Kingman coalescent.
(In fact, with $n=4$, this ratio captures \emph{all} of the distortion in the SFS.)
As with the larger sample size (\fig{fig:fpmi_vs_tajimasd}), mulitple mergers generate larger distortions in $\fpmi$ than population growth does, accounting for the distortions in the SFS.
Moreover, the results for two different models of growth coincide.
This suggests that for the Kingman coalescent with growth, $\fpmi_{12} \approx f(\eta_1/\eta_2)$ for some function $f$.
